Data Exploration: IUCN and AquaMaps

### 2015-11-17 data_explore.Rmd
updating figures with theme

library(ggplot2)
library(maps)
library(readr)
library(raster)
library(data.table)
library(tidyr)
library(dplyr)

@@

# Data files
#iucn spp cells WITHOUT extinct locations of IUCN species
iucn_spp_cells <- read_csv(file.path(dir_anx, 'explore_am_v_iucn/iucn_spp_cells_both.csv'), 
                           col_types = 'cdddd', progress = TRUE) %>%
  as.data.frame() %>%
  filter(presence != 5)

@@

#species list 
spp_list_base <- read.csv('shiny/data/spp_am_v_iucn.csv', stringsAsFactors = FALSE)

#data for raster (used to plot species)

loiczid_raster_file  <- 'shiny/data/loiczid_raster.grd'
loiczid_raster       <- raster(loiczid_raster_file)
names(loiczid_raster) <- 'loiczid'
```

```{r species_map_function, echo = FALSE, warning = FALSE, message = FALSE}

@@

  return(spp_map)
}
```
***

@@

barchart_sp_groups <- ggplot(spp_df, aes(x = spp_group)) +
  geom_bar(stat = 'bin') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) + 
  coord_flip() +
  labs(x = 'Species Group', 
       y = 'Number of Species', 
       title = 'Number of species in both AquaMaps and IUCN datasets')

@@

###(2) Looking at the total difference in area (km2) for each species range map in both datasets

@@

scatter_am_iucn_area <- ggplot(spp_df, aes(x = iucn_area, y = am_area, color = iucn_category)) + 
  geom_point() + 
  geom_abline(slope = 1, intercept = 0, color = 'red') +
  labs(x = 'Area (km2) from IUCN dataset', 
       y = 'Area (km2) from AquaMaps dataset', 
       title = 'Comparing total area of species range maps')

@@

scatter_am_iucn_area_log <- ggplot(spp_df, aes(iucn_log, am_log, color = iucn_category)) + 
  geom_point() + 
  geom_abline(slope = 1, intercept = 0, color = 'red') +
  labs(x = 'IUCN Area (log transformed)', 
       y = 'AquaMaps Area (log transformed)',
       title = 'Comparing area of range maps\n (all area values are km2 log transformed)', 
       fill = 'IUCN Category')

@@

scatter_am_iucn_alog_gp <- ggplot(spp_df, aes(iucn_log, am_log, color = iucn_category)) + 
  geom_point() +
  facet_wrap(~spp_group) + 
  geom_abline() + 
  labs(x = 'IUCN Area (log transformed)', 
       y = 'AquaMaps Area (log transformed)',
       title = 'Comparing differences in area across species groups', 
       fill = 'IUCN Category')

@@

spp <- 'Chromis alpha'
map <- get_spp_map(spp)

r_am_spp  <-  subs(loiczid_raster, 
                   am_map[ , c('loiczid', 'am_prob')], 
                   by = 'loiczid', 
                   which = 'am_prob', 
                   subsWithNA = TRUE)

r_am_spp[!is.na(r_am_spp)] <- 1

iucn_map <- map %>%
  mutate(iucn_pres = iucn_area > 0)

r_iucn_spp <- subs(loiczid_raster, 
                   iucn_map[ , c('loiczid', 'iucn_pres')], 
                   by = 'loiczid', 
                   which = 'iucn_pres', 
                   subsWithNA = TRUE)

### plot raster of species: IUCN map, AM map, world map, legend
plot(r_iucn_spp, 
     col       = 'mediumpurple', 
     useRaster = FALSE, 
     axes      = FALSE, 
     box       = FALSE, 
     legend    = FALSE)

plot(r_am_spp, 
     col       = 'orange', 
     main      = 'Yellow-speckled Chromis', 
     useRaster = FALSE, 
     add       = TRUE, 
     axes      = FALSE, 
     box       = FALSE, 
     legend    = FALSE)

map('world', 
    col    = 'gray95', 
    fill   = TRUE, 
    border = 'gray80', 
    add    = TRUE)

legend('topright', 
       legend = c('AquaMaps', 'IUCN'), 
       fill   = c('orange', 'mediumpurple'))

@@

hist_overlap_pct <- ggplot(overlap, aes(percent)) +
  geom_histogram(aes(y = ..count../1926), binwidth = 5, alpha = .8) +
  labs(x = '% total range that is shared b/w both datasets', 
       y = 'Proportion of Total Species')

@@

hist_overlap_pct_gp <- ggplot(overlap, aes(percent)) + 
  geom_histogram(binwidth = 5, alpha = .8) +
  facet_wrap(~spp_group) +
  labs(x = 'Percent of total range that is shared between both datasets', 
       y = 'Number of Total Species') + 
  theme(strip.text.x = element_text(size = 8))

@@

hist_sm_within_lg <- ggplot(spp_df4, aes(sm_perc, fill = sm_range)) + 
  geom_histogram(aes(y = ..count../sum(..count..)), binwidth = 4, alpha = .8) +
  scale_fill_discrete(name = 'Dataset with smaller range', labels = c('AquaMaps', 'IUCN')) + 
  labs(x = 'Percent of smaller range that falls within the larger range', 
       y = 'Proportion of all species (n = 1962)')

@@

scatter_area_vs_overlap <- ggplot(spp_list, aes(sm_perc, total_area, color = sm_range)) + 
  geom_point() + 
  geom_abline() + 
  labs(y = 'Total Area (km2)', 
       x = 'Percent overlap smaller range in larger', 
       title = 'Size of species range vs. percent overlap of \nsmaller range with larger')

@@

### Inspect each quadrant to find representative species
quadrant1_spp <- spp_list %>%
  filter(area_ratio > 90 & sm_perc > 90)
quadrant2_spp <- spp_list %>%
  filter(area_ratio > 25 & area_ratio < 30) %>%
  filter(sm_perc > 90)
quadrant4_spp <- spp_list %>%
  filter(sm_perc > 25 & sm_perc < 30) %>%
  filter(area_ratio > 90)
quadrant3_spp <- spp_list %>%
  filter(area_ratio < 25 & sm_perc < 25)

spp_quads <- c('Kajikia albida', 'Galaxea cryptoramosa', 'Acanthurus nigroris', 'Metanephrops australiensis', 'Conus magnificus')

spp_list_label <- spp_list %>%
  filter(sciname %in% spp_quads)

scatter_arearatio_vs_overlap <- ggplot(spp_list, aes(area_ratio, sm_perc, color = sm_range)) + 
  geom_point() + 
  geom_abline() + 
  geom_point(data = spp_list_label, color = 'black', size = 4) + 
  geom_text(data = spp_list_label, 
            aes(label = sciname), 
            color = 'black',
            size = 4,
            vjust = 1.5) +
  labs(x = 'Range area ratio (smaller:larger) * 100%', 
       y = 'Percent overlap smaller range in larger', 
       title = 'Ratio of species range vs. percent overlap of \nsmaller range with larger')

@@

for (spp in spp_quads) { # spp = 'Kajikia albida'
  map <- get_spp_map(spp)

r_am_spp  <-  subs(loiczid_raster, 
                   am_map[ , c('loiczid', 'am_prob')], 
                   by = 'loiczid', 
                   which = 'am_prob', 
                   subsWithNA = TRUE)

r_am_spp[!is.na(r_am_spp)] <- 1

iucn_map <- map %>%
  mutate(iucn_pres = iucn_area > 0)

r_iucn_spp <- subs(loiczid_raster, 
                   iucn_map[ , c('loiczid', 'iucn_pres')], 
                   by = 'loiczid', 
                   which = 'iucn_pres', 
                   subsWithNA = TRUE)

spp_pts <- as.data.frame(rasterToPoints(r_iucn_spp)) %>%
  full_join(as.data.frame(rasterToPoints(r_am_spp)),
            by = c('x', 'y')) %>%

spp_plot <- ggplot(spp_pts, aes(x = x, y = y)) +
  geom_raster(aes(fill = iucn_pres), alpha = .5) +
  geom_raster(aes(fill = am_prob), alpha = .5) + 
  borders('world', color='gray40', fill='gray45') + # create a layer of borders

### plot raster of species: IUCN map, AM map, world map, legend
plot(r_iucn_spp, 
     col       = '#ff6666', 
     main      = spp,
     alpha     = 1,
     useRaster = FALSE, 
     axes      = FALSE, 
     box       = FALSE, 
     legend    = FALSE)

plot(r_am_spp, 
     col       = '#6666ff', 
     alpha     = .4,
     useRaster = FALSE, 
     add       = TRUE, 
     axes      = FALSE, 
     box       = FALSE, 
     legend    = FALSE)

map('world', 
    col    = 'gray95', 
    fill   = TRUE, 
    border = 'gray80', 
    add    = TRUE)

legend('topright', 
       legend = c('AquaMaps', 'IUCN'), 
       fill   = c('#6666ff', '#ff6666'))
}

@@

scatter_a_vs_overlap_small <- ggplot(spp_small_df, aes(total_area, sm_perc, color = sm_range)) + 
  geom_point() + 
  labs(x = 'Total Area (km2)', 
       y = 'Percent overlap smaller range in larger', 
       title = 'Size of species range vs. percent overlap of \nsmaller range with larger')

@@

scatter_logA_vs_overlap <- ggplot(spp_list, aes(log_total_area, sm_perc, color = sm_range)) + 
  geom_point() + 
  labs(x = 'Log(total Area)', 
       y = 'Percent overlap smaller range in larger', 
       title = 'Size of species range (log) vs. percent overlap of \nsmaller range with larger')

@@

  qtile_plot <- ggplot(area_clip, aes(total_area, sm_perc, color = sm_range)) + 
    geom_point() + 
    labs(x = 'Total Area (km2)', 
         y = 'Percent overlap smaller range in larger', 
         title = sprintf('Size of species range (quartile %s) vs. percent overlap of \nsmaller range with larger', i))

@@

# get spp_cells raw data for aquamaps

dir_N <- c('Windows' = '//neptune.nceas.ucsb.edu/data_edit',
           'Linux'   = '/var/data/ohi')[[ Sys.info()[['sysname']] ]]

dir_aquamaps <- file.path(dir_N, 'git-annex/globalprep/SPP_ICO/raw/aquamaps_2014')

spp_cells <- fread(input = file.path(dir_aquamaps, 'tables/ohi_hcaf_species_native.csv'), verbose = FALSE, header = TRUE, 
                  colClasses = c('NULL', 'character', NA, 'numeric', 'NULL', 'NULL')) #1 min 49 sec

@@

spp       <- fread(input = file.path(dir_aquamaps, 'tables/ohi_speciesoccursum.csv'), 
                  verbose = FALSE, header = TRUE) %>%
              mutate(scientific = paste(Genus, Species, sep = ' '))
spp <- as.data.frame(spp)

@@

#create dataframe that gives me all species, their total number of cells, total number of cells >= 0.4 prob, differnece between the two and threat
spp_df3 <- spp_cells %>%
  group_by(species_id) %>%
  summarize(cells_0.4 = sum(probability >= 0.4),
            cells_t = sum(probability >= 0)) %>%
  mutate(prop = cells_0.4 / cells_t) %>%
  left_join(spp, by = c('species_id' = 'SPECIESID')) %>%
  select(scientific, species_id, cells_0.4, cells_t, prop, iucn_code)

hist_area_vs_threshold <- ggplot(spp_df3, aes(prop, fill = iucn_code)) + 
  geom_histogram(aes(y = ..count../sum(..count..)), binwidth = 0.05, alpha = .7) + 
  labs(x = 'Proportion of range area included when selecting a threshold of 0.4', y = 'Proportion of all aquamaps species (17,348 total)')

>>>>

library(ggplot2)
library(maps)
library(readr)
library(raster)
library(data.table)
library(tidyr)
library(dplyr)
library(stringr)

@@

# Data files
#iucn spp cells WITHOUT extinct locations of IUCN subpopulations
iucn_spp_cells <- read_csv(file.path(dir_anx, 'explore_am_v_iucn/iucn_spp_cells_both.csv'), 
                           col_types = 'cdddd', progress = TRUE) %>%
  as.data.frame() %>%
  filter(presence != 5) ### removes extinct subpopulations

@@

### species info list - scinames, IUCN species ID, AM species ID, category etc
spp_list_base <- read.csv('shiny/data/spp_am_v_iucn.csv', stringsAsFactors = FALSE)

### LOICZID cell ID numbers - used in plotting rasters
loiczid_raster_file  <- 'shiny/data/loiczid_raster.grd'
loiczid_raster       <- raster(loiczid_raster_file)
names(loiczid_raster) <- 'loiczid'
```

``` {r ggplot_theme_setups, echo = FALSE}

### generic theme for all plots
ggtheme_plot <- theme(axis.ticks = element_blank(),
                    text = element_text(family = 'Helvetica', color = 'gray30', size = 8),
                    plot.title = element_text(size = rel(1.5), hjust = 0, face = 'bold'),
                    legend.position = 'right')

# ggtheme_map <- theme(axis.ticks = element_blank(), axis.text = element_blank(),
#                     text = element_text(family = 'Helvetica', color = 'gray30', size = 8),
#                     plot.title = element_text(size = rel(1.5), hjust = 0, face = 'bold'),
#                     legend.position = 'right')

### theme for species range maps - based on generic plot theme
ggtheme_map <- ggtheme_plot +
  theme(axis.text = element_blank())

``` {r species_map_functions, echo = FALSE, warning = FALSE, message = FALSE}

@@

  return(spp_map)
}

### Plot Species Map function
### uses ggplot to create a formatted map of species ranges, including
### Aquamaps, IUCN, and overlapped ranges; returns the plot object.
plot_rangemap <- function(spp) {
  map <- get_spp_map(spp)

  r_am_spp  <-  subs(loiczid_raster, 
                     am_map[ , c('loiczid', 'am_prob')], 
                     by = 'loiczid', 
                     which = 'am_prob', 
                     subsWithNA = TRUE)
  r_am_spp[!is.na(r_am_spp)] <- 1

  iucn_map <- map %>%
    mutate(iucn_pres = iucn_area > 0)
  r_iucn_spp <- subs(loiczid_raster, 
                     iucn_map[ , c('loiczid', 'iucn_pres')], 
                     by = 'loiczid', 
                     which = 'iucn_pres', 
                     subsWithNA = TRUE)

  spp_pts <- as.data.frame(rasterToPoints(r_iucn_spp)) %>%
    full_join(as.data.frame(rasterToPoints(r_am_spp)),
              by = c('x', 'y')) %>%
    mutate(presence = ifelse(am_prob & is.na(iucn_pres), 'am',
                             ifelse(iucn_pres & is.na(am_prob), 'iucn',
                                    ifelse(iucn_pres & am_prob, 'both', NA))))

  spp_plot <- ggplot(spp_pts, aes(x = x, y = y)) +
    ggtheme_map + 
    geom_raster(aes(fill = presence), alpha = .5) +
    scale_fill_manual(values = c('iucn' = 'blue', 'am' = 'red', 'both' = 'mediumorchid')) +
    borders('world', color='gray40', fill='gray45', size = .1) +  # create a layer of borders
    labs(title = spp, x = NULL, y = NULL) 
}

```
***

@@

barchart_sp_groups <- ggplot(spp_df, aes(x = spp_group)) +
  ggtheme_plot + 
  geom_bar(stat = 'bin') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) + 
  coord_flip() +
  labs(x = 'Species Group', 
       y = 'Number of Species', 
       title = 'Number of species in both AquaMaps and IUCN datasets')

@@

###(2) Looking at the total difference in area (km^2) for each species range map in both datasets

@@

scatter_am_iucn_area <- ggplot(spp_df, aes(x = iucn_area, y = am_area, color = iucn_category)) + 
  ggtheme_plot + 
  geom_point() + 
  geom_abline(slope = 1, intercept = 0, color = 'red') +
  labs(x = 'Area (km^2) from IUCN dataset', 
       y = 'Area (km^2) from AquaMaps dataset', 
       title = 'Comparing total area of species range maps')

@@

scatter_am_iucn_area_log <- ggplot(spp_df, aes(iucn_log, am_log, color = iucn_category)) + 
  ggtheme_plot + 
  geom_point() + 
  geom_abline(slope = 1, intercept = 0, color = 'red') +
  labs(x = 'IUCN Area (log transformed)', 
       y = 'AquaMaps Area (log transformed)',
       title = 'Comparing area of range maps\n (all area values are km^2 log transformed)', 
       fill = 'IUCN Category')

@@

scatter_am_iucn_alog_gp <- ggplot(spp_df, aes(iucn_log, am_log, color = iucn_category)) + 
  ggtheme_plot + 
  geom_point() +
  facet_wrap(~spp_group) + 
  geom_abline() + 
  labs(x = 'IUCN Area (log transformed)', 
       y = 'AquaMaps Area (log transformed)',
       title = 'Comparing differences in area across species groups', 
       fill = 'IUCN Category')

@@

spp <- 'Chromis alpha'

spp_map <- plot_rangemap(spp)

print(spp_map)
ggsave(plot = spp_map,
       filename = sprintf('figures/rangemap_%s.png', spp),
       width = 8, height = 4, dpi = 200)
```

@@

hist_overlap_pct <- ggplot(overlap, aes(percent)) +
  ggtheme_plot + 
  geom_histogram(aes(y = ..count../1926), binwidth = 5, alpha = .8) +
  labs(title = 'Percent of total range shared between datasets',
       x = '% total range that is shared b/w both datasets', 
       y = 'Proportion of Total Species')

@@

hist_overlap_pct_gp <- ggplot(overlap, aes(percent)) + 
  ggtheme_plot + 
  geom_histogram(binwidth = 5, alpha = .8) +
  facet_wrap(~spp_group) +
  labs(title = 'Percent of total range shared, by species group',
       x = 'Percent of total range that is shared between both datasets', 
       y = 'Number of Total Species') + 
  theme(strip.text.x = element_text(size = 8))

@@

hist_sm_within_lg <- ggplot(spp_df4, aes(sm_perc, fill = sm_range)) + 
  ggtheme_plot + 
  geom_histogram(aes(y = ..count../sum(..count..)), binwidth = 4, alpha = .8) +
  scale_fill_discrete(name = 'Dataset with smaller range', labels = c('AquaMaps', 'IUCN')) + 
  labs(title = 'Percent of smaller range within larger range',
       x = 'Percent of smaller range that falls within the larger range', 
       y = sprintf('Proportion of all species (n = %s)', nrow(spp_df4)))

@@

scatter_area_vs_overlap <- ggplot(spp_list, aes(sm_perc, total_area, color = sm_range)) + 
  ggtheme_plot + 
  geom_point() + 
  geom_abline() + 
  labs(y = 'Total Area (km^2)', 
       x = 'Percent overlap smaller range in larger', 
       title = 'Size of species range vs. percent overlap of \nsmaller range with larger')

@@

### Inspect each quadrant to find representative species
# quadrant1_spp <- spp_list %>%
#   filter(area_ratio > 90 & sm_perc > 90)
# quadrant2_spp <- spp_list %>%
#   filter(area_ratio > 25 & area_ratio < 30) %>%
#   filter(sm_perc > 90)
# quadrant4_spp <- spp_list %>%
#   filter(sm_perc > 25 & sm_perc < 30) %>%
#   filter(area_ratio > 90)
# quadrant3_spp <- spp_list %>%
#   filter(area_ratio < 25 & sm_perc < 25)

spp_quads <- c('q1' = 'Kajikia albida', 
               'q2' = 'Galaxea cryptoramosa', 
               'q3' = 'Acanthurus nigroris', 
               'q3a' = 'Metanephrops australiensis', 
               'q4' = 'Conus magnificus')

spp_list_label <- spp_list %>%
  filter(sciname %in% spp_quads) %>%
  separate(sciname, c('g', 's'), sep = ' ', remove = FALSE) %>%
  mutate(sciname_label = paste(str_sub(g, 1, 1), '. ', s, sep = '')) %>%
  select(-g, -s)

scatter_arearatio_vs_overlap <- ggplot(spp_list, aes(area_ratio, sm_perc, color = sm_range)) + 
  ggtheme_plot + 
  geom_point() + 
  geom_abline() + 
  geom_point(data = spp_list_label, color = 'grey30') + 
  geom_text(data = spp_list_label, 
            aes(label = sciname_label), 
            color = 'grey30',
            fontface = 3,
            size = 2,
            hjust = .25,
            vjust = 1.5) +
  labs(x = 'Range area ratio (smaller:larger) * 100%', 
       y = 'Percent overlap smaller range in larger', 
       title = 'Ratio of species range vs. percent overlap of \nsmaller range with larger')

@@

for (i in 1:length(spp_quads)) { # i = 1
  spp <- spp_quads[i]
  spp_quad <- names(spp_quads)[i]

  spp_plot <- plot_rangemap(spp)
  print(spp_plot)

  ggsave(filename = sprintf('figures/rangemap_%s_%s.png', spp_quad, str_replace(tolower(spp), ' ', '_')),
         plot = spp_plot, 
         width = 8, height = 4, dpi = 200)
}

@@

scatter_a_vs_overlap_small <- ggplot(spp_small_df, aes(total_area, sm_perc, color = sm_range)) + 
  ggtheme_plot + 
  geom_point() + 
  labs(x = 'Total Area (km^2)', 
       y = 'Percent overlap smaller range in larger', 
       title = 'Size of species range vs. percent overlap of \nsmaller range with larger')

@@

scatter_logA_vs_overlap <- ggplot(spp_list, aes(log_total_area, sm_perc, color = sm_range)) + 
  ggtheme_plot + 
  geom_point() + 
  labs(x = 'Log(total Area)', 
       y = 'Percent overlap smaller range in larger', 
       title = 'Size of species range (log) vs. percent overlap of \nsmaller range with larger')

@@

  qtile_plot <- ggplot(area_clip, aes(total_area, sm_perc, color = sm_range)) + 
    ggtheme_plot + 
    geom_point() + 
    labs(x = 'Total Area (km^2)', 
         y = 'Percent overlap smaller range in larger', 
         title = sprintf('Size of species range (quartile %s) vs. percent overlap of \nsmaller range with larger', i))

@@

# get spp_cells raw data for aquamaps
dir_aquamaps <- file.path(dir_N, 'git-annex/globalprep/SPP_ICO/raw/aquamaps_2014')

spp_cells <- read_csv(file.path(dir_aquamaps, 'tables/ohi_hcaf_species_native.csv'), 
                      col_types = '_ccd__', progress = TRUE)

# spp_cells <- fread(input = file.path(dir_aquamaps, 'tables/ohi_hcaf_species_native.csv'), verbose = FALSE, header = TRUE, 
#                   colClasses = c('NULL', 'character', NA, 'numeric', 'NULL', 'NULL')) #1 min 49 sec

@@

spp <- fread(input = file.path(dir_aquamaps, 'tables/ohi_speciesoccursum.csv'), 
             verbose = FALSE, header = TRUE) %>%
  mutate(sciname = paste(Genus, Species, sep = ' '))
spp <- as.data.frame(spp)

@@

### create dataframe that gives me all species, their total number of cells, 
### total number of cells >= 0.4 prob, difference between the two and threat
spp_df3 <- spp_cells %>%
  group_by(species_id) %>%
  summarize(cells_0.4 = sum(probability >= 0.4),
            cells_t = sum(probability >= 0)) %>%
  mutate(prop = cells_0.4 / cells_t) %>%
  left_join(spp, by = c('species_id' = 'SPECIESID')) %>%
  select(sciname, species_id, cells_0.4, cells_t, prop, iucn_code)

hist_area_vs_threshold <- ggplot(spp_df3, aes(prop, fill = iucn_code)) + 
  ggtheme_plot + 
  geom_histogram(aes(y = ..count../sum(..count..)), binwidth = 0.05, alpha = .7) + 
  labs(x = 'Proportion of range area included when selecting a threshold of 0.4', y = 'Proportion of all aquamaps species (17,348 total)')
